---
title: "ExploratoryDataAnalysis"
output: html_notebook
editor_options: 
  chunk_output_type: console
chunk_output_type: inline
---

I have the data in front of me. What i am gonna do?


1. What is the goal?  
Our goal is to understand the data with descriptive statistics and visualization techniques

2. Tidy Data Process

3. Determining and Setting Variable Types

4. Summary statistics: Demonstrating the Basic Structure of the Data Set

5. Exploratory Data Analysis

  Exploratory Data Analysis refers to the critical process of performing initial investigations on    data so as to discover patterns,to spot anomalies,to test hypothesis and to check assumptions with the help of summary statistics and graphical representations.


```{r}
library(tidyverse)
```

## Variable Transformations and  setting data types 
```{r}
df <- mpg  # mpg is a dataset from R library
df$class <- factor(df$class)
glimpse(df)
# displ variable class is dbl.
df$disp_new <- ifelse(df$displ > 2, "one", "zero") # create a new variable and it is gonna be with only 2 values. If it is smaller bigger than 2, it will be 1. But if  it is smaller than 2, it will be 0. Then we need to set this variable as factor for our future use.
df$disp_new <- factor(df$disp_new) # convert "disp_new" variable to factor


```


## First look to the dataset
```{r}
colnames(mpg)
rownames(mpg)
nrow(mpg)
ncol(mpg)
head(mpg)
str(mpg)
glimpse(mpg)
psych::headTail(mpg)
summary(df)
fivenum(df$displ)
levels(df$class)
```


## Quick look to continious and categoric variables
```{r}
library(funModeling)
profiling_num(df) # mean' std_dev, variation cofficient...
plot_num(df)
freq(df)
```


## For more detail about continious and categoric variables
```{r}
library(psych)
describe(mpg$displ)

library(Hmisc)
Hmisc::describe(mpg$displ)

library(pastecs)
stat.desc(mpg)
mpg

df <- select(mpg, trans, cty, cyl, hwy)
describeBy(df, df$trans)
```


# Visualizing Distributions


## Bar Plot (visualisation of categorical variables)
```{r}

df <- diamonds #Here, I use diamonds dataset from R library
glimpse(df) # see the datatypes 

ggplot(df, aes(cut)) +
  geom_bar()

ggplot(df, aes(cut, fill = color)) +
  geom_bar(position = position_dodge()) + 
  ggtitle("Explore Diamonds data set") +
  xlab("Diamond quality") + 
  ylab("Observation Frequencies")

```



## Histogram and Density
```{r}

glimpse(df)
ggplot(df, aes(price)) +
  geom_histogram(binwidth = 1000)

ggplot(df, aes(price)) +
  geom_density()

ggplot(df, aes(price)) +
  geom_histogram(aes(y = ..density..)) +
  geom_density(alpha = .3, fill = "orange") 

ggplot(df, aes(price)) +
  geom_histogram() + 
  facet_grid(cut ~. )

```


## Ridgeline
```{r}
install.packages("ggridges")
library(ggridges)

ggplot(df, aes(x = price, y = cut, fill = cut)) +
  geom_density_ridges() + 
  theme_ridges() + 
  theme(legend.position = "none")

```



## Interactive Histogram
```{r}

library(plotly)
#Let's creat a normal distributed dataframe and have a look at it's interactive histogram graphic.
g <- plot_ly(x = rnorm(500), 
             type = "histogram") 
g

# When you interactively want to see 2 different variable in one graphic, use this function 
g <- plot_ly(x = rnorm(500), 
             opacity = 0.6, 
             type = "histogram") %>%  #this is first variable
  add_trace(x = rnorm(500) + 1) %>%  # this is second variable
  layout(barmode =  "overlay")
g

```



## Visualing Frequency of multiple elements 
```{r}
ggplot(df, aes(price, 
               y = ..density..)) +
  geom_density(aes(colour = cut), 
               binwidth = 500)

ggplot(df, aes(price, y = ..density..)) +
  geom_freqpoly(aes(colour = cut), binwidth = 500)

```



## Box Plot (use for visualization of continuos variables and detecting outliars)
```{r}

ggplot(df, aes(x = cut, 
               y = price,
               fill = cut)) +
         geom_boxplot()

```


## Violin Graph 
```{r}
ggplot(df, aes(x = cut, 
               y = price, 
               fill = cut)) +
         geom_violin(alpha = 0.6)

```



# Analysis of Correlations between variables

## Scatter Plot  (for 2 variables) ( At the section below "correlation matrixes"; you can see the  correlation for 3 or more variables)

```{r}

library(tidyverse)

glimpse(iris)
df <- iris 

# see the correlation between sepal.length and sepal.width
ggplot(df, aes(Sepal.Length, Sepal.Width)) +
  geom_point(size = 2,  # shows observation points
             shape = 21, 
             stroke = 1,
             color = "black",
             fill = "orange") +
  geom_smooth(color = "red")  # shows correlation line

# add a categorical variable (species) to the same graph above
ggplot(df, aes(Sepal.Length, 
               Sepal.Width, 
               color = Species, 
               shape = Species )) + 
  geom_point(size = 6, alpha = 0.6) 
  
# add color and size to petal.length
ggplot(df, aes(Sepal.Length, 
               Sepal.Width, 
               color = Petal.Length, 
               size = Petal.Length )) +
  geom_point(alpha = 0.6) 


```




### Print graph of observations with 3 variables
```{r}

df <- mtcars

ggplot(df, aes(x = wt, y = mpg, fill = cyl, color = "Yellow")) +
  geom_point() + 
  geom_label(label = rownames(df),  
            nudge_x = 0.25, 
            nudge_y = 0.2) +
  geom_smooth(method = lm, se = FALSE)


```




## HeatMap (Use for understand the correlation between categorical variable and continious variable)
```{r}
# to use the heatmap function we should convert the dataframe to matrix
# we should delete categorical variables if exists to use this function

df <- as.matrix(mtcars) 
str(df)
heatmap(df, scale = "column") 
heatmap(df, Colv = NA, Rowv = NA, scale = "column")

```




## Correlation matrixes  (understand the correlation between more than 2 variables)
```{r}
library(GGally)

df <- mtcars[ , c(1,3:6)]

cor(df)
cor.test(df$mpg, df$wt)
plot(df)

ggcorr(df)
ggpairs(df)

```



## Time series visualization
```{r}

df <- economics

ggplot(df, aes(date, pop))



ggplot(df, aes(date, pop)) +
  geom_line(color = "Orange", size =1)

ggplot(df, aes(date, pop)) +
  geom_line(aes( size = (unemploy/pop)))

ggplot(df, aes(date, psavert)) +
  geom_line() +
  stat_smooth()
```


## TreeMap

Treemaps are a data visualization technique that uses internal square-rectangular structures to visualize hierarchical data sets.


Each branch of the tree is assigned a category in the data as a rectangle, and its sub-branches are assigned sub-categories of that category. 
On the bottom of the leaves are the continuous variable values belonging to these categories are shown, so we need a continuous variable.

### Simple TreeMap (without sub-groups)
```{r}

library(treemap)

df <- data.frame(
  groups = c("group_1", "group_2","group_3"),
  values = c(10, 90, 60)
  
)
 
df

treemap(df, index = "groups", vSize = "values", type ="index")
```


### 2.Creating treemap with sub-groups
```{r}




# The code below creates a dataframe 
df <- data.frame(
  groups = c(rep("group_1", 4), #rep("group_1", 4) function creates  4 element with group_1 group
               rep("group_2", 2),
               rep("group_3", 3)),
  
  sub_groups = paste("sub_group", c(1,2,3,4,1,2,1,2,3), sep = "_"),
  
  values = c(13,5,22,12,11,4,2,5,6)
  
)
df # see the dataframe


# see the groups and sub-groups at the dataframe with Treemap function
treemap(df, index = c("groups","sub_groups"), vSize ="values")


# add vusual options to the same graph
treemap(df, index = c("groups","sub_groups"), 
        vSize = "values", 
        type ="index",
        fontsize.labels = c(20,13),
        fontcolor.labels = c("white","black"),
        bg.labels = c("transparent"),
        align.labels = list(
          c("center", "center"),
          c("right","bottom")
          
        ),
        border.col = c("black","white"),
        border.lwds = c(5,2)
        )

# load diamonds dataset and see the treemap graph for 'cut', 'color' and 'price' variables
df<- diamonds
df <- select(diamonds, cut, color, price)

treemap(df, index = c("cut","color"), 
        vSize = "price", 
        type ="index",
        fontsize.labels = c(20,13),
        fontcolor.labels = c("white","black"),
        bg.labels = c("transparent"),
        align.labels = list(
          c("center", "center"),
          c("right","bottom")
          
        ),
        border.col = c("black","white"),
        border.lwds = c(5,2)
        )







```
